\documentclass[11pt]{letter}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\signature{Author One\\ on behalf of all authors}
\address{Institute/Department\\ National Taiwan University\\ Taipei, Taiwan\\ email@ntu.edu.tw}
\begin{document}
\begin{letter}{The Editors\\ \emph{Briefings in Bioinformatics}\\ Oxford University Press}
\opening{Dear Editors,}

We are pleased to submit our manuscript, ``\emph{How should genomic foundation
models be evaluated for metagenomic taxonomic classification? A controlled
benchmark of pre-training, tokenization, and data scaling},'' for consideration
as a \textbf{Problem-solving Protocol} in \emph{Briefings in Bioinformatics}.

Genomic foundation models (GFMs) are increasingly proposed for metagenomic
tasks, but the community lacks clear, controlled guidance on \emph{which} design
choice actually determines their performance for short-read taxonomic
classification. Published comparisons typically vary model size, data, and
tokenization together, leaving practitioners unable to attribute gains. Our
manuscript addresses this methodological question directly, in the
protocol-and-recommendation style that fits the journal's readership rather than
as a single new-method paper.

We believe the work suits \emph{Briefings in Bioinformatics} for three reasons.
First, its contribution is a \emph{benchmarking protocol and a set of practical
recommendations}, not a single tool---each experiment isolates one design factor
(data scale, pre-training, tokenization) so the resulting guidance is directly
attributable and actionable. Second, the central finding is broadly useful and
somewhat counter-intuitive: tokenization, not backbone pre-training or data
scale, sets the ceiling for short-read taxonomy, and a non-neural 13-mer
composition model can outperform a 498M-parameter pre-trained model. Third, we
provide a dual read-level/sample-level evaluation and a coverage-matched
comparison with Kraken\,2 (both raw and with Bracken re-estimation) that yields
concrete, use-case-specific recommendations. Notably, a careful baseline shows
that Kraken\,2's apparent abundance weakness is a raw-count artefact that Bracken
corrects ($r=0.997$); we therefore place the neural models' genuine value in
read-level accuracy and out-of-reference behaviour rather than over-claiming a
sample-level abundance win. The manuscript is written to help readers decide whether to fine-tune
an existing GFM, train a task-specific $k$-mer transformer, use a classical
exact-$k$-mer classifier, or design a new microbial foundation model---and it
provides a reusable evaluation protocol for doing so.

We are transparent about the scope and its main limitation: the high closed-set
accuracy reflects near-complete reference coverage rather than novel-genome
generalization. We therefore delimit the study explicitly as closed-set
short-read classification and identify out-of-genome generalization as the
central next benchmark, rather than presenting the 13-mer result as evidence of
novel-genome discovery. The manuscript already includes a Bracken-corrected
Kraken\,2 abundance baseline and a two-replicate real mock-community probe
(ZymoBIOMICS D6331), which shows that the classical pipeline's simulated
in-database dominance does not transfer to real reads---reinforcing that the
binding open problem is the closed-set-to-real gap, not a missing baseline.

This manuscript is original, is not under consideration elsewhere, and all
authors approve its submission. We disclose in the manuscript that an AI
assistant was used to help draft and edit the text and organize results; all
scientific claims and reported numbers were verified by the authors. We declare
no competing interests. We would be glad to suggest suitable reviewers with
expertise in genomic language models and metagenomic classification upon
request.

\closing{Sincerely,}
\end{letter}
\end{document}
