%=======================================================================
%  Briefings in Bioinformatics -- Problem-solving Protocol
%  Official OUP class: oup-authoring-template (contemporary, numbered refs)
%
%  "How should genomic foundation models be evaluated for metagenomic
%   taxonomic classification?"
%
%  Compiles with pdflatex + bibtex on any full TeX Live / Overleaf.
%  The OUP class + .bst files are bundled in this repo, so no extra
%  installation is required.  Build order:  pdflatex -> bibtex -> pdflatex x2
%  (or: latexmk -pdf main.tex).
%=======================================================================
\documentclass[unnumsec,webpdf,contemporary,large,numbered]{oup-authoring-template}

% --- extra packages not already provided by the class ---
% (class already loads: graphicx, caption, amsmath, amssymb, array, xcolor,
%  natbib, multirow, url, hyperref, tikz)
% NB: tables use the class's tabular*/\toprule/\botrule house style, not
% tabularx (tabularx is incompatible with the class's rule macros).
\usetikzlibrary{arrows.meta,positioning,calc}

% Enable natbib sort&compress so multi-citations render as [8-14], not
% [8, 9, 10, 11, 12, 13, 14]. (The class loads natbib itself, so we set the
% switches directly rather than via a package option.)
\makeatletter
\def\NAT@sort{\@ne}\def\NAT@cmprs{\@ne}
\makeatother

\graphicspath{{figures/}}

% The OUP class defaults the running-head society logo to a PNAS Nexus image
% that is not shipped with the bundle; blank it out (documented in the class).
\def\societylogo{}

% colour used by the Figure 1 schematic
\definecolor{bibblue}{RGB}{0,63,114}

% For peer review you may enable line numbers (uncomment):
%\usepackage[mathlines,switch]{lineno}\linenumbers

\begin{document}

\journaltitle{Briefings in Bioinformatics}
\DOI{}
\copyrightyear{2026}
\pubyear{2026}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Problem-solving Protocol}

\firstpage{1}

\title[GFMs for metagenomic taxonomic classification]{How should genomic
foundation models be evaluated for metagenomic taxonomic classification?
A controlled benchmark of pre-training, tokenization, and data scaling}

\author[1,$\ast$]{First Author\ORCID{0000-0000-0000-0000}}
\author[1]{Second Author}

\address[1]{\orgdiv{Institute/Department}, \orgname{National Taiwan University},
\orgaddress{\street{Street}, \postcode{Postcode}, \state{Taipei},
\country{Taiwan}}}

\corresp[$\ast$]{Corresponding author. \href{mailto:email@ntu.edu.tw}{email@ntu.edu.tw}}

%\received{Date}{0}{Year}
%\revised{Date}{0}{Year}
%\accepted{Date}{0}{Year}

\abstract{\textbf{Background:} Genomic foundation models (GFMs) have shown
promise across many genomic tasks, yet their utility for metagenomic
short-read taxonomic classification---where reads are $\sim$150\,bp and
per-read taxonomic signal is sparse---remains poorly characterized, and it is
unclear which design axis (pre-training, tokenization, or data scale) actually
governs performance.
\textbf{Methods:} We present a controlled benchmark that varies one factor at a
time while holding architecture, data, and evaluation constant. Using reads
simulated from a unified human-gut genome catalogue across 120 genera and
1{,}535 species, we ablate data scale (500K--250M reads), pre-training
(a 498M-parameter pre-trained backbone versus a randomly initialized one),
tokenization (6/12/13-mer, overlapping versus non-overlapping), and evaluation
protocol (read-level accuracy, sample-level abundance estimation, and binary
detection), and compare against the alignment-free classifier Kraken\,2.
\textbf{Results:} Overlapping 13-mer tokenization, not backbone pre-training,
determines the performance ceiling: an overlapping-13-mer model trained from
scratch reaches 87.4\% genus Top-1 accuracy versus 67.1\% for a pre-trained
6-mer model on identical 50M-read data. The 6-mer model saturates beyond 50M
reads ($+0.22$ percentage points to 250M) while the 13-mer model keeps scaling
(87.4\% to 98.7\%); the 6-mer model fails even to \emph{fit} its training set
beyond $\sim$66\%, indicating a representational rather than an optimization
ceiling. With tokenization fixed at 6-mer, pre-training contributes $+13.2$
percentage points. At the sample level, under controlled synthetic compositions
a 67\%-accurate read classifier attains Pearson $r=0.99$ for genus abundance;
raw Kraken\,2 is deflated by its abstention ($r=0.82$) but Bracken re-estimation
recovers it ($r=0.997$), so on in-database data a properly configured
Kraken\,2$+$Bracken pipeline matches the neural models on abundance and remains
the stronger detector---the distinct value of GFMs is read-level accuracy under a
good tokenizer and behaviour outside the reference database, not sample-level
abundance. On a real mock community (two D6331 replicates), however, all methods
(neural and Kraken\,2$+$Bracken) fall to $r\approx0.4$--$0.6$ with overlapping
uncertainty and no clear winner, exposing a large closed-set-to-real gap that is
the binding open problem.
\textbf{Conclusion:} Tokenization design is the critical, and currently
under-appreciated, factor for GFM-based metagenomic classification. We translate
these findings into use-case-specific recommendations and argue that future
microbial foundation models should be pre-trained with overlapping long-$k$
tokenization.}

\keywords{genomic foundation models, metagenomics, taxonomic classification,
tokenization, benchmarking, short-read sequencing}

\boxedtext{Key Points}{
\begin{itemize}
\item Data scaling is tokenizer-dependent: increasing training data from 50M to
250M reads yields almost no gain for a non-overlapping 6-mer model
($+0.22$ percentage points) but an 11-point gain for an overlapping 13-mer model
($87.4\%\!\rightarrow\!98.7\%$).
\item Tokenization, not backbone pre-training, sets the performance ceiling: an
overlapping 13-mer model trained from scratch reaches 87.4\% genus Top-1 versus
67.1\% for a 498M-parameter pre-trained model under identical data.
\item Pre-training still helps when tokenization is held fixed ($+13.2$
percentage points over a randomly initialized backbone with the same 6-mer
input), so the two factors are separable and both matter.
\item Read-level accuracy and sample-level utility diverge: a 67\%-accurate read
classifier reaches Pearson $r=0.99$ for genus abundance on simulated
in-database samples, but that ranking does not transfer to a real mock
community (two D6331 replicates: $r\approx0.4$--$0.6$ for all methods, including
Kraken\,2$+$Bracken, with overlapping uncertainty at $n{=}11$ genera).
\item We give practical, use-case-specific recommendations and argue that future
microbial foundation models should be pre-trained with overlapping long-$k$
tokenization.
\end{itemize}}

\maketitle

\input{sections/01_introduction}
\input{sections/02_challenges}
\input{sections/03_benchmark_design}
\input{sections/04_data_scaling}
\input{sections/05_pretraining_vs_tokenization}
\input{sections/06_read_vs_sample}
\input{sections/07_recommendations}
\input{sections/08_limitations}
\input{sections/09_endmatter}

% Use unsrtnat (numbered, in order of appearance = Vancouver/BIB style).
% NB: do NOT use the bundled oup-plain.bst here -- it emits bare \bibitem{key}
% entries, which makes the class flip natbib back to author-year mode and
% produces empty "[ , ]" citations. unsrtnat writes labelled \bibitem[..]{key}
% so the numeric citations resolve correctly.
\bibliographystyle{unsrtnat}
\bibliography{references}

\end{document}
