\section{Practical recommendations}
\label{sec:recommendations}

The controlled results above translate into concrete guidance. We separate
recommendations for practitioners choosing or applying a tool today from
recommendations for those designing the next generation of models. Box~1 first
distils the benchmark itself into a reusable protocol; Box~2 is a decision guide
for choosing a method.

\begin{boxtext}[A protocol for evaluating GFMs in short-read taxonomy]
\begin{enumerate}[1.]\setlength{\itemsep}{1pt}
\item Fix the taxonomy level (genus/species) and read length up front.
\item Build both read-disjoint \emph{and} genome-disjoint splits; report
closed-set and out-of-reference results separately.
\item Report nominal read counts \emph{and} unique-read counts (short-read
simulation duplicates reads; scaling claims must not be duplicate scaling).
\item Vary one axis at a time---pre-training, tokenizer, data scale---holding the
architecture and evaluation fixed.
\item Evaluate read-level accuracy, sample-level abundance, and presence/absence
detection \emph{separately}; they do not track each other.
\item Compare against both a raw exact-$k$-mer classifier (e.g.\ Kraken\,2) and an
abundance-corrected pipeline (e.g.\ Kraken\,2$+$Bracken).
\item State the strand-symmetry protocol (e.g.\ reverse-complement TTA) and apply
it consistently across models.
\item Release configs, exact commands, per-read predictions, and sample
composition files.
\end{enumerate}
\end{boxtext}

\subsection{For practitioners}

\begin{table*}[t]
\centering
\caption{\textbf{Use-case-specific recommendations.} ``long-$k$ model'' denotes
an overlapping long-$k$-mer neural classifier; ``6-mer GFM'' denotes a
pre-trained NT-v2-style model with non-overlapping 6-mer tokenization.}
\label{tab:reco}
\begin{tabular}{@{}p{0.30\textwidth}p{0.63\textwidth}@{}}
\toprule
Use case & Recommendation \\
\midrule
Read-level genus classification &
Prefer an overlapping long-$k$ model (87\% vs 67\%); a 6-mer GFM is a distant
second. \\[2pt]
Sample-level abundance (composition) &
For simulated in-database samples, Kraken\,2$+$Bracken is excellent ($r=0.997$)
and matches the best neural models ($r\!\approx\!0.99$); do not expect a neural
model to beat a Bracken-corrected pipeline on clean closed-set data. Compare
against raw counts \emph{and} Bracken. On a real mock community
(Section~\ref{sec:mock}), however, that ranking does not transfer: all methods
fall to $r\approx0.4$--$0.6$ with no clear winner, and always-labelling neural
models (including a 6-mer GFM) are competitive on Bray--Curtis and detection.
Treat simulated abundance rankings as provisional until real-community checks.
\\[2pt]
Presence/absence detection, esp.\ low-abundance taxa &
On simulated in-database data, prefer Kraken\,2 (or another exact-$k$-mer tool);
current 6-mer neural detectors are weak there. On real mocks the picture is
mixed (neural models can show higher sensitivity when Kraken\,2 abstains
heavily). Apply a calibrated abundance threshold above the $\sim$0.3\% noise
floor. \\[2pt]
Species-level classification &
6-mer GFMs are insufficient; use a long-$k$ tokenizer or a hierarchical router,
or an exact-$k$-mer tool when the target is in-database. \\[2pt]
Handling class imbalance &
Do \emph{not} aggressively genus-balance if you report abundance: it lifts macro
metrics but harms per-read accuracy and abundance fidelity ($r\!:0.99\!\to\!0.86$). \\[2pt]
Choosing a GFM &
Do not select on backbone size alone; the tokenizer decides whether taxonomic
signal survives. Ask what $k$ and what stride the model uses. \\
\botrule
\end{tabular}
\end{table*}

\begin{boxtext}[Decision guide]
\begin{itemize}\setlength{\itemsep}{1pt}
\item \textbf{Low-abundance presence/absence?} Prefer an exact-$k$-mer pipeline
(Kraken\,2/Bracken, Centrifuge) on well-covered references; current 6-mer neural
detectors are weak there. On real communities with heavy abstention, re-check
against a neural baseline.
\item \textbf{Known-genome, closed-set read classification?} Use an overlapping
long-$k$ neural model (87\% vs 67\% at genus).
\item \textbf{Only a 6-mer GFM available?} Do not expect data scaling to lift the
ceiling; the tokenizer, not the data, is the limit.
\item \textbf{Community abundance (in-database, simulated)?} Use
Kraken\,2$+$Bracken ($r=0.997$); a neural model is comparable but not better.
\item \textbf{Community abundance (real / incomplete reference)?} Do not trust
the simulated ranking: on a real mock, neural and Kraken\,2$+$Bracken are
competitive with no clear winner (Section~\ref{sec:mock}).
\item \textbf{Novel-genome discovery?} The current closed-set evidence is
insufficient; out-of-genome validation is required first.
\end{itemize}
\end{boxtext}

\subsection{For model builders}

The single highest-leverage design choice is tokenization. Future microbial
foundation models intended for short-read taxonomy should be \emph{pre-trained}
with overlapping long-$k$ tokenization (e.g.\ 12--13-mers), combining the
discriminative power that overlapping long $k$-mers provide with the
representational quality that pre-training adds---the two independent gains we
quantified ($+20$ pp from tokenization, $+13$ pp from pre-training). Because a
naive long-$k$ vocabulary is enormous (a 13-mer vocabulary implies a
multi-billion-parameter embedding table; the ``$\sim$5M-parameter''
MetaTransformer figure counts only the encoder and understates true cost),
practical realizations will need vocabulary factorization, hashing, or hybrid
tokenization---an open engineering problem we regard as the key next step.
Pre-training corpora should draw on the hundreds of thousands of microbial
genomes now catalogued \citep{Parks2022,Almeida2019} rather than
eukaryote-dominated data.
