%=======================================================================
%  ENDMATTER STATEMENTS  (required by Briefings in Bioinformatics)
%  Section titles follow the OUP authoring-template conventions.
%=======================================================================
\section{Conflicts of interest}
The authors declare that they have no competing interests.

\section{Funding}
[Funding sources and grant numbers to be inserted.] Computation was supported by
the National Center for High-performance Computing (NCHC), Taiwan.

\section{Data availability}
The reference genome catalogue is available from its original source
\citep{Almeida2019}. Simulated reads, trained model checkpoints, per-read
prediction files, and the sample-level evaluation outputs are deposited at
[repository name] at [URL/DOI to be inserted on acceptance]. The read-simulation,
training, and evaluation code is available at
\url{https://github.com/m2lab-ntu/gfm-classifier}.

\section{Code availability}
All training, inference, and evaluation scripts, together with model
configuration files and the exact training commands used for each experiment,
are provided in the repository above and summarized in the Supplementary
Material to enable reproduction.

\section{Author contributions statement}
Using the CRediT taxonomy (authors identified by initials):
X.X. conceived and designed the study, implemented the software, ran the
experiments, analysed the results, and wrote the original draft;
Y.Y. supervised the work, acquired resources and funding, and reviewed and
edited the manuscript. All authors read and approved the final manuscript.
% Replace initials and roles with the actual author list before submission.

\section{Acknowledgments}
The authors thank [colleagues/reviewers] for helpful discussions. During the
preparation of this work an AI coding/writing assistant was used to help draft
and edit the manuscript \LaTeX{} and to organize experimental results. All
experimental design, analysis, interpretation, and final scientific claims are
the authors' own; the authors verified all reported numbers against the
underlying results and take full responsibility for the content. No AI tool is
listed as an author.
